---
title: "Werewolves and Factions and Katniss, Oh My!"
author: Olivia DeMeo and Lilly Farah
date: May 9, 2019 at 11:55 PM
output:
  html_document:
    code_folding: hide
---
Young Adult novels and series are a genre of literature that entertains a large population of the greater public. From vampires and werewolves, to dystopian settings, each novel tells a story that transports the reader into a new world.

What way to make something more realistic than a visual representation of the words within the pages? Movie adaptations of popular YA series are particularly interesting due to the multiple movies within a series. The real question, though, is are they worth the money put into them? How well do they do with audiences? In order to answer these questions we will comparte each movie's budget and total gross in the United States. By determining the profits we see the success of [*The Twilight Saga*](https://en.wikipedia.org/wiki/The_Twilight_Saga_%28film_series%29), [*The Hunger Games*](https://en.wikipedia.org/wiki/The_Hunger_Games_%28film_series%29) series, and the [*Divergent*](https://en.wikipedia.org/wiki/The_Divergent_Series) series. Furthermore, the success of the previous movies within the series is a determining factor for the following movie’s budget.

```{r, message = FALSE, echo = TRUE}
library(tidyverse)
library(ggrepel)
library(RMySQL)
db <- dbConnect(MySQL(), 
                host = "scidb.smith.edu", 
                user = "mth292", 
                password = "RememberPi", 
                dbname = "imdb")
knitr::opts_chunk$set(connection = db, max.print = 20)
```

```{sql, output.var = "budget"}
SELECT title AS movie, production_year, REPLACE(SUBSTRING(info FROM 2), ',', '') AS budgetnum
FROM movie_info
RIGHT JOIN title ON movie_info.movie_id = title.id
WHERE movie_info.info_type_id = '105'
AND (movie_id = '2969942' or movie_id = '4368658' or movie_id = '4368665' or movie_id = '4368666'
or movie_id = '3518460' or movie_id = '3754229' or movie_id = '3270236' or movie_id = '4368646'
or movie_id = '4490027' or movie_id = '4436841' or movie_id = '4436842' or movie_id = '4436846' or movie_id = '4436851')
```

```{sql, output.var = "budget", message = FALSE}
SELECT title AS movie, production_year, REPLACE(SUBSTRING(info FROM 2), ',', '') AS budgetnum
FROM movie_info
RIGHT JOIN title ON movie_info.movie_id = title.id
WHERE movie_info.info_type_id = '105'
AND (movie_id = '2969942' or movie_id = '4368658' or movie_id = '4368665' or movie_id = '4368666'
or movie_id = '3518460' or movie_id = '3754229' or movie_id = '3270236' or movie_id = '4368646'
or movie_id = '4490027' or movie_id = '4436841' or movie_id = '4436842' or movie_id = '4436846' or movie_id = '4436851')
```

First, we looked at each movie and its given budget. The visualization below maps each movie by the year it was produced and its corresponding gross value.

```{sql, output.var = "gross", message = FALSE}
SELECT title AS movie, production_year, REPLACE(REPLACE(SUBSTRING(info FROM 2 FOR 11), ',', ''), ' ', '') AS grossing
FROM movie_info
RIGHT JOIN title ON movie_info.movie_id = title.id
WHERE movie_info.info_type_id = '107'
AND (movie_id = '2969942' or movie_id = '4368658' or movie_id = '4368665' or movie_id = '4368666'
or movie_id = '3518460' or movie_id = '3754229' or movie_id = '3270236' or movie_id = '4368646'
or movie_id = '4490027' or movie_id = '4436841' or movie_id = '4436842' or movie_id = '4436846' or movie_id = '4436851')
AND (info LIKE '%(USA)' OR info LIKE '%(22 May 2016)' OR info LIKE '%(21 October 2010)' 
OR info LIKE '%(29 June 2010)' OR info LIKE '%$192,769,854%' OR info LIKE '%$281,287,133%' OR info LIKE '%(3 March 2013)')
```
```{r, message = FALSE, warning=FALSE}
budget <- budget %>% 
  mutate(budgetnum = as.numeric(budgetnum) / 1000000) %>%
  mutate(series = ifelse(grepl("Twilight", movie), "Twilight", ifelse(grepl("Hunger", movie), "The Hunger Games", "Divergent")))

ggplot(budget, aes(x = production_year, y = budgetnum, label = movie)) +
  geom_point() +
  geom_smooth(color = "magenta") +
  geom_label_repel(alpha = .7, aes(color = series)) +
  labs(title = "Budget of Popular YA Series", y = "Budget (in millions)", x = "Production Year", caption = "Source: IMDb", color = "Series")
```

The general trend upward may be caused by the success of previous movies within the genre. Also, within each series, the budgets increased for each successive film. From this information, it seems as though all of the movies became more successful as there were more sequels produced. *Twilight* started the boom of YA movie adaptations with a budget of under $50 million. The success of the first movie of the franchise correlates with the higher budget of the second movie, and an increase in budget for each movie made after. The last movie in the series, *The Twilight Saga: Breaking Dawn - Part 2*, has a budget that is higher than the first *Hunger Games* movie, which was made in the same year. This agrees with our intuition because *Twilight* was an established franchise, while *The Hunger Games* franchise was just testing the waters of the film industry. With this in mind, we can also see that *The Hunger Games* had a larger budget than the first three *Twilight* movies. The investors may have seen the success of *The Twilight Saga* and realized that this genre of movie does well with the general public. Trends also show that the *Hunger Games* movies had the largest budgets, which may correlate to their immense popularity as novels by [Suzanne Collins](https://en.wikipedia.org/wiki/Suzanne_Collins). The *Divergent* series had budgets within the same range as the *Hunger Games*, which is expected considering the success of the previous two YA movie franchises and the increased interest and production of the YA film series.


Next, we compared the trends in gross for each movie and each series. The following data visualization maps each movie by the year it was produced and its corresponding gross value.

```{r, message = FALSE, warning=FALSE}
gross <- mutate(gross, grossing = as.numeric(grossing) / 1000000) %>%
  mutate(series = ifelse(grepl("Twilight", movie), "Twilight", ifelse(grepl("Hunger", movie), "The Hunger Games", "Divergent")))

ggplot(gross, aes(x = production_year, y = grossing, label = movie)) +
  geom_point() +
  geom_smooth(color = "magenta") +
  geom_label_repel(alpha = .7, aes(color = series)) +
  labs(title = "Grossing of Popular YA Series", y = "Gross (in millions)", x = "Production Year", caption = "Source: IMDb", color = "Series")
```

This was not what we expected to see as a trend. The most obvious feature of the graph is the steep decline in gross for the later movies produced. In the years between 2008 and 2013, the gross trend line has an increasing slope, which represents the increase in popularity and interest in the movie franchises and the YA genre of movies. After 2013, there is a sharp decline in gross numbers, with *Allegiant* grossing less than $100 million. The *Hunger Games* series did well with the first two movies, *The Hunger Games* and *The Hunger Games: Catching Fire*, but the two parts of the final movie were not as successful and show the start of the genre’s decline in popularity and prosperity. The *Divergent* series was still produced, though the hopes of large profits were not as extensive as those of the previous series. The decrease in gross of the *Divergent* movies joined the trend of decreasing popularity of the YA movie adaptations.


Despite the prevalence and success of these young adult movies in the past, there haven’t been any movies released of the same caliber in the past years. Given the trend seen with the these franchises, especially with the *Divergent* series, this may be because the movies starting grossing less and less, even if the budgets were higher. The *Divergent* series, while a popular book series, was cut short, and the last movie in the series was never made. This may have been a sign to other filmmakers that this genre was not as profitable as others. The few blockbusters marketed at a young adult audience in the past few years may be due to the *Divergent* series’ underperformance and the decreasing grossing of *The Hunger Games* series.

From these two comparisons, we have coined a new term that refers to an era of time, called the “YA Movie Boom”, which lasted a total of about 8 years, with a peak in 2012. It will be interesting to see how this genre continues in the future, especially with the recent increase in the Marvel franchise, which has surpassed these series’ at the box office. What will be the next big movie genre to climb the charts? 
^[[Github](https://github.com/lillyfarah/SDS192-MiniProject4)]